---
title: "pubtatordb"
author: "Zachary Colburn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{pubtatordb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Overview

[PubTator](https://www.ncbi.nlm.nih.gov/CBBresearch/Lu/Demo/PubTator/) is an NCBI product that contains detailed annotations of abstracts found on PubMed. This makes it a very useful research tool. While PubTator does provide an API, the use of an API is inconvenient for high-throughput analyses and also requires a guaranteed internet connection. Querying a local PubTator database is better suited for high-throughput analyses. The package pubtatordb makes it easy to quickly start using a local copy of PubTator's data.


## Installation

You can install the released version of pubtatordb from [CRAN](https://CRAN.R-project.org) with:

```{r, eval=FALSE}
install.packages("pubtatordb")
```


The version on GitHub can be downloaded using the devtools package with:

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("mamc-dci/pubtatordb")
```


## Example


Load the package.
```{r message=FALSE, warning=FALSE}
# Load the package.
library(pubtatordb)
```


After loading the package, database setup and querying can be accomplished in four steps.

- Data download
- Database setup
- Connect to the database
- Query the database


After the user manually creates a folder to store the data, the user can define the path to that folder and then download the data to that location:
```{r, eval=FALSE}
# Download the data.
# Use the full path. Writing to the temp directory is not recommended.
download_dir <- tempdir()
download_pt(download_dir)
```


After defining the path to the download directory created above, the database can be created with:
```{r, eval=FALSE}
# Define the data directory, a subdirectory of the above directory.
pubtator_path <- file.path(download_dir, "PubTator")

# Create the database.
pt_to_sql(
  pubtator_path,
  skip_behavior = FALSE,
  remove_behavior = TRUE
)
```


If the .gz files from PubTator have already been extracted, their extraction can be skipped with the **skip_behavior** argument. After their insertion into the database, both the .gz and uncompressed files can be removed using the **remove_behavior** argument.


A connection can be created to the database using **pt_connector**. Note that this is a wrapper for the **dbConnect** function of the DBI package.
```{r, eval=FALSE}
# Create a connection to the database.
db_con <- pt_connector(pubtator_path)
```


Querying the data is accomplished using the **pt_select** function. The first five rows of the gene table can be selected with:
```{r, eval=FALSE}
# Query the data.
pt_select(
  db_con,
  "gene",
  columns = NULL,
  keys = NULL,
  keytype = NULL,
  limit = 5
)
```


The first five results for PMIDs in which the genes with ENTREZ IDs 7356 or 4199 were mentioned can be selected with:
```{r, eval=FALSE}
# Query the data.
pt_select(
  db_con,
  "gene",
  columns = c("PMID", "ENTREZID"),
  keys = c("7356", "4199"),
  keytype = "ENTREZID",
  limit = 5
)
```


## Other tables
PubTator has several datasets. The names of tables in the database can be obtained with:


```{r, eval=FALSE}
pt_tables(db_con)
```


The column names for a particular table can be accessed with:
```{r, eval=FALSE}
pt_columns(db_con, "species")
```


## Note
The citation information for PubTator can be found on the [PubTator](https://www.ncbi.nlm.nih.gov/CBBresearch/Lu/Demo/PubTator/) website or with:
```{r}
pubtator_citations()
```


## Disclaimer
The views expressed are those of the author(s) and do not reflect the official policy of the Department of the Army, the Department of Defense or the U.S. Government.
